### 2017-01-25 manuscript.rmd
Update manuscript.rmd
code formatting

```{r setup, cache=TRUE, message=F, echo=FALSE, warning=F}
knitr::opts_chunk$set(echo = TRUE)
```

@@

^4^ Sars Institute, Bergen, Norway

@@

Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw] analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm]. Using pairwise comparisons (Fig. 1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.

@@

> Figure 1 | Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies across species rely on pairwise comparisons, where traits of each gene are compared to traits of every other gene. This leads to many more comparisons than unique observations, making each comparison non-independent. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.

@@

#### Original pairwise analyses

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution. Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortholog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough expression study to date.

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of Brawand *et al.* 2011 [@Brawand:2011du], one of the studies they considered. This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their reanalyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. For example, it does not convey which tissue expression is specific to. Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.

@@

	expression_file_names %<>% paste(kmrr_directory, ., sep="")

	gene_tree_name %<>% paste(kmrr_directory, ., sep="")

	# These are from the KMRR code
	calibration_times = data.frame(
		age.min =
			c(20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86), 
		clade = c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
							"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
							"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
							"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
							"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"),
		stringsAsFactors=FALSE
	)

@@

			# Create calibration matrix for speciation nodes
			calibration = 
				nhx@data[ !is.tip.nhx( nhx ) ,  ] %>%
				filter( D=="N" ) %>%
				left_join( calibration_times, c( "label" = "clade" ) ) %>%
				select( node, age.min ) %>%
				mutate( age.max = age.min ) %>% 
				mutate( soft.bounds = NA )

			tree = try( ape::chronos( nhx@phylo, calibration=calibration, model="correlated" ) )
			if( "phylo" %in% class(tree) ){
				class( tree ) = "phylo"
				nhx@phylo = tree
				return( nhx )
			}
			else{
				return( NA )
			}
		}

@@

```{r load_expression, echo=FALSE, message=F, warning=F}
	expression = lapply(
		expression_file_names, read.table, stringsAsFactors=FALSE, header=TRUE
	) %>%
	bind_rows()
```

@@

	gene_trees = mclapply(
		lines,
		function( x ){
			tree_tc = textConnection( x )
			tree <- treeio::read.nhx( tree_tc )
			close( tree_tc )
			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
			return( tree )
		}, mc.cores=cores
	)
```

```{r add_expression_to_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
	gene_trees_annotated = mclapply(
		gene_trees,
		function( tree ){
			tree@data %<>% left_join( expression, by = c("G" = "Ensembl.Gene.ID") )
			return( tree )
		}, mc.cores=cores
	)
```

```{r prep_trees, echo=FALSE, cache=TRUE, message=F, warning=F}
	drop_empty_tips = function( nhx ) {
	  to_drop = which( is.na( nhx@data[ 1:length(nhx@phylo$tip.label), ]$Tau ) )
	  remaining = length(nhx@phylo$tip.label) - length(to_drop)

@@

	all_nodes = lapply( 
	      gene_trees_pic, 
	  		function( nhx ){
	  			tags = nhx@data
	  			tags$gene = digest( nhx )
	  			tags %<>% select( -(B) ) # B has inconsistent types, remove it
	  			return(tags)
	  		}
  		) %>%
		bind_rows()

@@

	nodes_contrast = 
		all_nodes %>%
		filter( ! is.na(pic) ) %>%
		mutate( pic=abs(pic) ) %>%
		filter( ! is.na(D) )

@@

	wilcox_test_result = 
		wilcox.test( 
			nodes_contrast %>% filter(D=="Y") %>% .$pic, 
			nodes_contrast %>% filter(D=="N") %>% .$pic, 
			alternative="greater" 
		)

@@

  g_age =
  nodes_contrast %>%
    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
		  geom_point( alpha=0.2 ) + 
		  xlab( "Node age (millions of years)" ) + 
		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
		  ylim(0,0.3) +
			xlim(0,1000) +
		  ggtitle( "(a)" ) +
		  theme_classic()

@@

	wilcox_filtered_age = 
		wilcox.test( 
			nodes_contrast_filtered_age %>% filter(D=="Y") %>% .$pic, 
			nodes_contrast_filtered_age %>% filter(D=="N") %>% .$pic, 
			alternative="greater" 
		)$p.value

@@

  nodes_contrast_filtered_var = 
    nodes_contrast %>% 
      filter ( var_exp >=var_exp_min & var_exp <=var_exp_max )

	wilcox_filtered_var = 
	wilcox.test( 
		nodes_contrast_filtered_var %>% filter(D=="Y") %>% .$pic, 
		nodes_contrast_filtered_var %>% filter(D=="N") %>% .$pic, 
		alternative="greater" 
	)$p.value

@@

		# Calculate the phenotypic difference between the tips with a different method as 
		# check on parsing
		D$difference_old = 
			apply(
				D,
				1,
				function(y) {
					tau_values = nhx@data$Tau[ c( as.integer(y['tip_a']), as.integer( y['tip_b'] ) ) ]
					max( tau_values ) - min( tau_values )
				}
			)

		if ( ! all ( near( D$difference, D$difference_old ) ) ){
			stop( "Inconsistent difference calculations" )
		}

		# For each row, bet the most recent common ancestor of the pairwise tip combination
		D$mrca = 
			apply(
				D, 
				1, 
				function(y) ape::getMRCA( 
					nhx@phylo, 
					c( as.integer(y['tip_a']), as.integer(y['tip_b'] ))
				)
			) %>%
			as.integer()

@@

	ortholog_distance_mean = 
		pairwise_summary %>% 
		filter(D=="N") %>% 
		.[["distance"]] %>% 
		mean()

	paralog_distance_mean = 
		pairwise_summary %>% 
		filter(D=="Y") %>% 
		.[["distance"]] %>% 
		mean()

@@

	g_density_filtered_age =   
		nodes_contrast_filtered_age %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
      geom_freqpoly( binwidth=.001, position="identity" ) +
      xlab("Phylogenetic independent \ncontrasts (PIC)") +
			ylab("Density") +
			xlim(0,0.075) + 
			ggtitle( "(b)" ) +
			annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
			theme_classic()

	# Examine by expected variance (branch length)
  g_var_exp =
    nodes_contrast %>%
      ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
      geom_point( alpha=0.2 ) + 
      xlab( "Expected variance, \n depends mostly on branch length" ) + 
      ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
      ylim(0,0.3) +
			xlim(0,2000) +
      ggtitle( "(c)" ) +
      theme_classic()	
      # theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

@@

  g_density_filtered_var =   
    nodes_contrast_filtered_var  %>% 
    ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
      geom_freqpoly( binwidth=.001, position="identity" ) + 
      ggtitle( "(d)" ) +
      theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
      xlim(0,0.075) + 
			theme_classic() +
  		annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
      xlab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
  		ylab("Density")

@@

```{r pairwise_plot, echo=FALSE, cache=TRUE, message=F, warning=F}	
	pairwise_summary %>% 
		filter( ! is.na(D) ) %>%
  	ggplot(aes(x=distance, y=difference, color=D)) + 
  		geom_point(alpha=0.1) +
  		geom_smooth(method=lm, se=FALSE) +
			xlim( 0, 2000 )

>>>>

```{r setup, cache=TRUE, message=F, echo=FALSE, warning=F}

	knitr::opts_chunk$set( echo = TRUE )

@@

^4^ Sars International Centre for Marine Molecular Biology, University of Bergen, Bergen, Norway

@@

The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend of have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm]. Using pairwise comparisons (Fig. 1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.

@@

> Figure 1 | Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison non-independent. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.

@@

#### Original pairwise test of the ortholog conjecture

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that have not been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortholog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.

The KMRR study considered several publicly available datasets. Here we consider their conclusions in the context of one of these studies, Brawand *et al.* 2011 [@Brawand:2011du]. This dataset of gene expression for six organs across 10 species (nine mammals and a bird) is the best sampled in their analyses. Their expression summary statistic is Tau [@Yanai:2005dh, @KryuchkovaMostacci:2016in], an indicator of tissue specificity of gene expression. It can range from a value of 0, which indicates no specificity (*i.e.*, uniform expression across tissues), to a value of 1, which indicates high specificity (*i.e.*, expression in only one tissue). It is convenient in that it is a single number of defined range for each gene, though of course since the original expression is multidimensional this means much information is discarded. For example, it does not convey which tissue expression is specific to. Homologous genes could each have a tau value of 1, but one gene could have expression that is specific to the kidney and the other to the brain, for example.

@@

	expression_file_names %<>% paste( kmrr_directory, ., sep="" )

	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )

	# These are from the KMRR code
	calibration_times = data.frame(
		age.min =
			c( 20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86 ), 
		clade = 
			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
					"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
					"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
					"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
					"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"
			),
		stringsAsFactors=FALSE
	)

@@

			# Create calibration matrix for speciation nodes
			calibration = 
				nhx@data[ !is.tip.nhx( nhx ), ] %>%
				filter( D == "N" ) %>%
				left_join( calibration_times, c( "label" = "clade" ) ) %>%
				select( node, age.min ) %>%
				mutate( age.max = age.min ) %>% 
				mutate( soft.bounds = NA )

			tree = try( 
				ape::chronos( nhx@phylo, calibration=calibration, model="correlated" ) 
			)

			if( "phylo" %in% class( tree ) ){
				class( tree ) = "phylo"
				nhx@phylo = tree
				return( nhx )
			}
			else{
				return( NA )
			}
		}

@@

```{r load_expression, echo=FALSE, message=F, warning=F}

	expression = 
	lapply(
		expression_file_names, 
		read.table, 
		stringsAsFactors=FALSE, 
		header=TRUE
	) %>%
	bind_rows()

@@

	gene_trees = mclapply(
		lines,
		function( x ){
			tree_tc = textConnection( x )
			tree = treeio::read.nhx( tree_tc )
			close( tree_tc )
			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
			return( tree )
		}, mc.cores=cores
	)
```

```{r add_expression_to_trees, echo=FALSE, cache=TRUE, message=F, warning=F}

	gene_trees_annotated = mclapply(
		gene_trees,
		function( tree ){
			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )
			return( tree )
		}, 
		mc.cores=cores
	)

```{r prep_trees, echo=FALSE, cache=TRUE, message=F, warning=F}

	drop_empty_tips = function( nhx ) {
		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )
		remaining = length( nhx@phylo$tip.label ) - length( to_drop )

@@

	all_nodes = 
		lapply( 
			gene_trees_pic, 
			function( nhx ){
				tags = nhx@data
				tags$gene = digest( nhx )
				tags %<>% select( -( B ) ) # B has inconsistent types, remove it
				return( tags )
			}
		) %>%
		bind_rows()

@@

	nodes_contrast = 
		all_nodes %>%
		filter( ! is.na( pic ) ) %>%
		mutate( pic = abs( pic ) ) %>%
		filter( ! is.na( D ) )

@@

	wilcox_test_result = 
		wilcox.test( 
			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
			alternative="greater" 
		)

@@

	g_age =
		nodes_contrast %>%
		ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
			geom_point( alpha=0.2 ) + 
			xlab( "Node age (millions of years)" ) + 
			ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
			ylim( 0, 0.3 ) +
			xlim( 0, 1000 ) +
			ggtitle( "(a)" ) +
			theme_classic()

@@

	wilcox_filtered_age = 
		wilcox.test( 
			nodes_contrast_filtered_age %>% filter( D=="Y" ) %>% .$pic, 
			nodes_contrast_filtered_age %>% filter( D=="N" ) %>% .$pic, 
			alternative="greater" 
		)$p.value

@@

	nodes_contrast_filtered_var = 
		nodes_contrast %>% 
		filter ( ( var_exp >= var_exp_min ) & ( var_exp <= var_exp_max ) )

	wilcox_filtered_var = 
	wilcox.test( 
		nodes_contrast_filtered_var %>% filter( D=="Y" ) %>% .$pic, 
		nodes_contrast_filtered_var %>% filter( D=="N" ) %>% .$pic, 
		alternative="greater" 
	)$p.value

@@

		# For each row, bet the most recent common ancestor of the pairwise tip combination
		D$mrca = 
			apply(
				D, 
				1, 
				function( y ) ape::getMRCA( 
					nhx@phylo, 
					c( as.integer( y['tip_a'] ), as.integer( y['tip_b'] ) )
				)
			) %>%
			as.integer()

@@

	ortholog_distance_mean = 
		pairwise_summary %>% 
		filter( D=="N" ) %>% 
		.[["distance"]] %>% 
		mean()

	paralog_distance_mean = 
		pairwise_summary %>% 
		filter( D=="Y" ) %>% 
		.[["distance"]] %>% 
		mean()

@@

	g_density_filtered_age =
		nodes_contrast_filtered_age %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		ggtitle( "(b)" ) +
		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
		theme_classic()

	# Examine by expected variance (branch length)
	g_var_exp =
		nodes_contrast %>%
		ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
		geom_point( alpha=0.2 ) + 
		xlab( "Expected variance, \n depends mostly on branch length" ) + 
		ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
		ylim( 0, 0.3 ) +
		xlim( 0, 2000 ) +
		ggtitle( "(c)" ) +
		theme_classic()

@@

	g_density_filtered_var =
		nodes_contrast_filtered_var %>% 
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) + 
		ggtitle( "(d)" ) +
		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
		xlim( 0, 0.075 ) + 
		theme_classic() +
		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
		xlab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
		ylab( "Density" )

@@

```{r pairwise_plot, echo=FALSE, cache=TRUE, message=F, warning=F}	
	pairwise_summary %>% 
	filter( ! is.na( D ) ) %>%
	ggplot( aes( x=distance, y=difference, color=D ) ) + 
		geom_point( alpha=0.1 ) +
		geom_smooth( method=lm, se=FALSE ) +
		xlim( 0, 2000 )
